---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# covidscreen

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of {covidscreen} is to provides simple simulations for understanding
the effects of assurance testing/screening on a population. It aims to help
decision-makers determine whether and how to test for COVID-19 with limited
resources.

## Installation

You can install the current version of {covidscreen} with:

``` r
# install.packages("remotes")
remotes::install_github("jesse-smith/covidscreen")
```

## Thoughts

1. Add different % asymptomatic for vaccinated & unvaccinated
2. Assume Rt constant, rather than infection rate?
3. Use exponential distributions for time-based variables, rather than constants

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(covidscreen)
## basic example code
```

## Code of Conduct
  
Please note that the covidscreen project is released with a
[Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.
