---
title: "covidscreen: A Simple, Simulation-Based Tool for Assessing Organizational COVID-19 Testing Strategies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{covidscreen: A Simple, Simulation-Based Tool for Assessing Organizational COVID-19 Testing Strategies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(covidscreen)
```

# Abstract

covidscreen is an interactive tool designed to allow decision-makers to assess the efficacy of their testing programs. While the underlying model is intentionally simple and understandable, it can be tuned to a variety of local and organizational contexts.

# Introduction

As SARS-CoV-2 continues to spread and mutate, organizations must make difficult decisions about resource allocation toward a seemingly never-ending threat. Many organizations have deployed both regular COVID-19 testing strategies and vaccination requirements in an attempt to keep their members safe. Exactly what level of intervention is needed, though, is often unknown, leading to either frequently questioning of spending on these programs or blind reliance on a decision made with inadequate information.

While predicting the precise level of intervention suitable for an organization requires context-specific details and possibly advanced modeling, an estimate aligning with real-world data and suitable for driving appropriate action is easier to achieve. To this end, we develop a simple probabilistic model of COVID-19 testing strategies, and we wrap this model in an interface designed to help decision-makers determine the best testing strategy for their organization and risk tolerance.

# Results

# Methods

# Discussion
